---
title: "Populate MRE Object"
author: "Marc Noguera"
date: "2/16/2022"
output: html_document
---
```{r setup, include=T,echo=F,message=F, eval=T}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = here::here())

```

## Initial Analysis

Using the initial MRE object which has all results slots empty, we then populate these slots through pre-programmed functions that use project-specifci categorical and numerical variables to run statistical testing of different nature:

* Ordination:
* Differential Abundance:
* Correlation:
* Gene Richness:
* Biomarker Discovery:
* Gene Function:
* Other: Project specific analysis may be programmed from scratch or can take advantage from pre-run statistical testing. 

All analysis generate by default a set of result files that are stored into disk, generally pdf or csv that allow result exploration.
Once all te tests have been run, the MRE is populated with the correspoonding results and saved into disk/AWS.

```{r embedded populate slots, echo=T, include=T, message=T, eval=T}
require(phyloseq)
require(ggstatsplot)
require(ggplot2)
require(here)

# Load package from local copy
devtools::load_all("~/Documents/Work/Development/WMGSPipeline/")### Change to require/library/load local package
# require(metar)

# myMREObject<-readRDS("data/processed/mreObject.rds")
myMREObject<-readRDS("data/processed/mreObject.rds")

# # ## QC
# myMREObject <- myMREObject %>%
#    dada2_qc(save_files = T) ### Not keeping rarefaction result slot, too big.
# # # 
# # # ## Alpha Diversity
myMREObject <- myMREObject %>%
  dada2_rarefaction() ### dada2_rarefaction tries to use GeneRichness --> Needs to be corrected

# myMREObject <- myMREObject %>%
#     dada2_richness()

### BARPLOTS
myMREObject<-myMREObject %>%
  dada2_barplots(tax_level="Phylum") %>% 
   dada2_barplots(tax_level="Genus",top_n = 50)
### Heatmap
myMREObject<-myMREObject %>%
  dada2_heatmap(tax_level = "Genus")
### NMDS

myMREObject<-myMREObject%>%
    dada2_nmds(tax_level="Genus")

myMREObject<-myMREObject %>%
    dada2_lefse() 

saveRDS(myMREObject,paste0(here::here(),"/data/processed/mreObject.rds"))
# aws.s3::s3saveRDS(myMREObject,bucket=bucket,object = paste0(bucketDir,"/MREObject.rds"))

```


```{r embedded populate slots Baseline, echo=T, include=T, message=T, eval=T}
require(phyloseq)
require(ggstatsplot)
require(ggplot2)
require(here)

# Load package from local copy
devtools::load_all("~/Documents/Work/Development/WMGSPipeline/")### Change to require/library/load local package

myMREObject<-readRDS("data/internal/baseline/mreObject.rds")

# ## QC
# myMREObject <- myMREObject %>%
#    dada2_qc(save_files = T) ### Not keeping rarefaction result slot, too big.
# # 
# # # ## Alpha Diversity
myMREObject <- myMREObject %>%
  dada2_rarefaction() ### dada2_rarefaction tries to use GeneRichness --> Needs to be corrected
# 
# myMREObject <- myMREObject %>%
#     dada2_richness()


### BARPLOTS
myMREObject<-myMREObject %>%
  dada2_barplots(tax_level="Phylum") %>% 
   dada2_barplots(tax_level="Genus",top_n = 50)
myMREObject<-myMREObject %>%
  dada2_heatmap(tax_level = "Genus")
myMREObject<-myMREObject%>%
    dada2_nmds(tax_level="Genus")
myMREObject<-myMREObject %>%
    dada2_lefse() 
saveRDS(myMREObject,paste0(here::here(),"/data/internal/baseline/mreObject.rds"))
# aws.s3::s3saveRDS(myMREObject,bucket=bucket,object = paste0(bucketDir,"/MREObject.rds"))

```


```{r embedded populate slots Treatment, echo=T, include=T, message=T, eval=T}
require(phyloseq)
require(ggstatsplot)
require(ggplot2)
require(here)

# Load package from local copy
devtools::load_all("~/Documents/Work/Development/WMGSPipeline/")### Change to require/library/load local package

myMREObject<-readRDS("data/internal/treatment/mreObject.rds")

# ## QC
# myMREObject <- myMREObject %>%
#    dada2_qc(save_files = T) ### Not keeping rarefaction result slot, too big.
# # 
# # # ## Alpha Diversity
myMREObject <- myMREObject %>%
  dada2_rarefaction() ### dada2_rarefaction tries to use GeneRichness --> Needs to be corrected

# myMREObject <- myMREObject %>%
#     dada2_richness()


### BARPLOTS
myMREObject<-myMREObject %>%
  dada2_barplots(tax_level="Phylum") %>% 
   dada2_barplots(tax_level="Genus",top_n = 50)
myMREObject<-myMREObject %>%
  dada2_heatmap(tax_level = "Genus")
myMREObject<-myMREObject%>%
    dada2_nmds(tax_level="Genus")
myMREObject<-myMREObject %>%
    dada2_lefse() 
saveRDS(myMREObject,paste0(here::here(),"/data/treatment/baseline/mreObject.rds"))
# aws.s3::s3saveRDS(myMREObject,bucket=bucket,object = paste0(bucketDir,"/MREObject.rds"))

```


```{r embedded populate slots Control, echo=T, include=T, message=T, eval=T}
require(phyloseq)
require(ggstatsplot)
require(ggplot2)
require(here)

# Load package from local copy
devtools::load_all("~/Documents/Work/Development/WMGSPipeline/")### Change to require/library/load local package

myMREObject<-readRDS("data/internal/placebo/mreObject.rds")

# ## QC
# myMREObject <- myMREObject %>%
#    dada2_qc(save_files = T) ### Not keeping rarefaction result slot, too big.
# 
# # ## Alpha Diversity
# myMREObject <- myMREObject %>%
#   dada2_rarefaction() ### dada2_rarefaction tries to use GeneRichness --> Needs to be corrected

# myMREObject <- myMREObject %>% 
#     dada2_richness()


### BARPLOTS
myMREObject<-myMREObject %>%
  dada2_barplots(tax_level="Phylum") %>% 
   dada2_barplots(tax_level="Genus",top_n = 50)
myMREObject<-myMREObject %>%
  dada2_heatmap(tax_level = "Genus")
myMREObject<-myMREObject%>%
    dada2_nmds(tax_level="Genus")
myMREObject<-myMREObject %>%
    dada2_lefse() 
saveRDS(myMREObject,paste0(here::here(),"/data/placebo/baseline/mreObject.rds"))
# aws.s3::s3saveRDS(myMREObject,bucket=bucket,object = paste0(bucketDir,"/MREObject.rds"))

```
