---
site: workflowr::wflow_site
title: "Microbiome association with Physical Activity"
author: "Marc Noguera-Julian, PhD. "
date: "`r format(Sys.time(), '%Y-%m-%d')`"
header-includes:
- \usepackage{fancyhdr}
- \usepackage{lipsum}
- \pagestyle{fancy}
- \fancyhead[CO,CE]{Microbioma intestinal en intervencion nutricional}
- \fancyfoot[LO,RE]{Marc Noguera-Julian, IrsiCaixa}
- \fancyhead[LE,RO]{\thepage}
- \fancyfoot[CO,CE]{}
- \usepackage[default]{sourcesanspro}
- \usepackage[T1]{fontenc}
- \usepackage{eso-pic,graphicx,transparent}
mainfont: Helvetica
output: html_document
---

```{r setup, include=TRUE, echo=F, message=FALSE, warning=FALSE, paged.print=FALSE, tidy=F}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_knit$set(root.dir = here::here())
rmd_env <- "devel"
```

## Physical Activity - Baseline

We will use the International Physical Activity Questionnaire  ([IPAQ](https://pubmed.ncbi.nlm.nih.gov/16925881/ PubMed Paper)) to assess whether participant phyisical activity shows any association with microbiome composition or derived parameters. 

### Alpha Diversity - Baseline

We will check, at baseline, if IPAQ shows any association with Alpha diversity measures (Chao1, Shannon, PD). Results indicate no correlation between any of the alpha-diversity parameters and MedDietScore.

```{r BL Alpha Diversity Chunk 1, echo=F, message=FALSE, warning=FALSE, paged.print=FALSE, tidy=F}
suppressWarnings(require(phyloseq))
suppressWarnings(require(phylosmith))
suppressWarnings(require(patchwork))
suppressWarnings(require(vegan))
suppressWarnings(require(ggstatsplot))
# Load package from local copy
suppressWarnings(devtools::load_all("~/Documents/Work/Development/WMGSPipeline/"))
ps_silva<-readRDS("data/cfu/DADA2/ps_silva.rds")
myMREObject<-readRDS("data/cfu/mreObject.rds")
myMREObject@wd<-"output/mre_data/cfu"

p0<-phyloseq::sample_data(ps_silva) %>% data.frame() %>% dplyr::filter(Timepoint=="BL") %>% 
  dplyr::rename(IPAQ=METActividadFisicabasaltotal) %>% 
  dplyr::select(IPAQ,chao1,diversity_shannon,PD) %>% 
  ggstatsplot::ggscatterstats(x=IPAQ,y=chao1,type="np",bf.message = NULL,title = "Chao1 vs IPAQ | BL")

p1 <- phyloseq::sample_data(ps_silva) %>% data.frame() %>% dplyr::filter(Timepoint=="BL") %>% 
  dplyr::rename(IPAQ=METActividadFisicabasaltotal) %>% 
  dplyr::select(IPAQ,chao1,diversity_shannon,PD) %>% 
  dplyr::rename(Shannon=diversity_shannon) %>% 
  ggstatsplot::ggscatterstats(x=IPAQ,y=Shannon,type="np",bf.message = NULL,title = "Shannon vs IPAQ | BL")

p2 <- phyloseq::sample_data(ps_silva) %>% data.frame() %>% dplyr::filter(Timepoint=="BL") %>% 
  dplyr::rename(IPAQ=METActividadFisicabasaltotal) %>% 
  dplyr::select(IPAQ,chao1,diversity_shannon,PD) %>% 
  dplyr::rename(Shannon=diversity_shannon) %>% 
  dplyr::rename(PhyloDist=PD) %>% 
  ggstatsplot::ggscatterstats(x=IPAQ,y=PhyloDist,type="np",bf.message = NULL,title = "PD vs IPAQ | BL")

p0
p1
p2

```

## Including Plots

You can also embed plots, for example:

```{r BL Ordination chunk 1, echo=F, message=FALSE, warning=FALSE, paged.print=FALSE, tidy=F}
suppressWarnings(require(phyloseq))
suppressWarnings(require(phylosmith))
suppressWarnings(require(patchwork))
suppressWarnings(require(vegan))
suppressWarnings(require(ggstatsplot))
# Load package from local copy
suppressWarnings(devtools::load_all("~/Documents/Work/Development/WMGSPipeline/"))
ps_silva<-readRDS("data/cfu/DADA2/ps_silva.rds")
myMREObject<-readRDS("data/cfu/mreObject.rds")
myMREObject@wd<-"output/mre_data/cfu"

### BL
BL_samples<-myMREObject %>% 
  metar::get_meta() %>% 
  dplyr::filter(Timepoint=="BL")

auxMREObject<-myMREObject %>%
  metar::filter_samples(.,BL_samples$SampleID) 

auxMREObject@metadata@categorical_vals<-tibble::tribble(~CategoricalVariable,~PaletteName,"METActividadFisicabasaltotal","Set1")

auxMREObject@taxa@dada2@phyloseq <- auxMREObject@taxa@dada2@phyloseq %>% 
  phyloseq::transform_sample_counts(., function(x) ((x/sum(x))))
auxMREObject<-metar::dada2_nmds(tax_level = "Genus", mre = auxMREObject,top_n = 50, save_files = F)

auxMREObject@taxa@dada2@nmds$Genus$top_50$numeric$METActividadFisicabasaltotal &
   theme(plot.title = element_text(size=12)) &
   theme(plot.subtitle = element_text(size=7))

```

## Intervention

### IPAQ Changes

We want to assess it there ar changes in physical activity after the intervention and if there are any differences between groups in order to assess the effectivity of the intervention. We can see that the median change in IPAQ score is close to 0 in both IG and CG, without significant differences between them.

```{r Intervention chunk 1, echo=F, message=FALSE, warning=FALSE, paged.print=FALSE, tidy=F}
suppressWarnings(require(phyloseq))
suppressWarnings(require(phylosmith))
suppressWarnings(require(patchwork))
suppressWarnings(require(vegan))
suppressWarnings(require(ggstatsplot))
# Load package from local copy
suppressWarnings(devtools::load_all("~/Documents/Work/Development/WMGSPipeline/"))
ps_silva<-readRDS("data/cfu/DADA2/ps_silva.rds")
myMREObject<-readRDS("data/cfu/mreObject.rds")
myMREObject@wd<-"output/mre_data/cfu"

phyloseq::sample_data(ps_silva) %>% data.frame() %>% 
  dplyr::rename(IPAQ=METActividadFisicabasaltotal) %>% 
  dplyr::select(Participante,Center,IPAQ,Timepoint) %>% 
  dplyr::group_by(Participante) %>% 
  tidyr::pivot_wider(values_from = IPAQ,names_from = Timepoint) %>% 
  dplyr::mutate(IPAQDiff=(Month6-BL)) %>% 
  dplyr::mutate(IPAQRelDiff=((Month6-BL)/(Month6+BL))) %>% 
  ggstatsplot::ggbetweenstats(x=Center,y=IPAQRelDiff,type="robust",bf.message = NULL,title="MDScore longitudinal changes vs Center")

```
