---
site: workflowr::wflow_site
title: "Intervention Analysis"
author: "Marc Noguera-Julian, PhD. "
date: "`r format(Sys.time(), '%Y-%m-%d')`"
header-includes:
- \usepackage{fancyhdr}
- \usepackage{lipsum}
- \pagestyle{fancy}
- \fancyhead[CO,CE]{Microbioma intestinal en intervencion nutricional}
- \fancyfoot[LO,RE]{Marc Noguera-Julian, IrsiCaixa}
- \fancyhead[LE,RO]{\thepage}
- \fancyfoot[CO,CE]{}
- \usepackage[default]{sourcesanspro}
- \usepackage[T1]{fontenc}
- \usepackage{eso-pic,graphicx,transparent}
mainfont: Helvetica
output: html_document
---

```{r setup, include=TRUE, echo=F, message=FALSE, warning=FALSE, paged.print=FALSE, tidy=F}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_knit$set(root.dir = here::here())
rmd_env <- "devel"
```


## Pairing Analysis

From Differential Abundance Analysis we have seen that there are a bunch of related ggenera that are overabundant in Month6 samples only in the IG. These genera happen to be all from the Lachnospiraceae family, except for Faecalibacterium.

Paired Boxplots clearly show that increase/decrease of these genera happens for only a subgroup of the participants. We are interested in confirming of discarding if this genus-specific increase/decrease subgroups of participants do overlap among genera.

```{r pairing analysis chunk 1, include=TRUE, echo=F, message=FALSE, warning=FALSE, paged.print=FALSE, tidy=T,eval=T}
suppressMessages(require(dplyr))
suppressMessages(require(phyloseq))
suppressMessages(require(ggplot2))
suppressMessages(require(patchwork))

set.seed(1234)
### Species Level

devtools::load_all("~/Documents/Work/Development/WMGSPipeline/")
myMREObject<-readRDS("data/cfu/mreObject.rds")

### Genus Level C01/IG
if(file.exists("output/Deliverables/DifferentialAbundance/maaslin/Volcano_Month6_vs_BL_IGsamples.rds")){
  fit_data<-readRDS("output/Deliverables/DifferentialAbundance/maaslin/Volcano_Month6_vs_BL_IGsamples.rds")
} 

sign_genus<-fit_data$results %>% 
                  dplyr::filter(qval<0.05) %>% 
                  dplyr::pull(feature)

ps_df<-myMREObject@taxa@dada2@phyloseq %>% 
  phyloseq::tax_glom(.,taxrank = "Genus",NArm = F) %>% 
  microbiome::transform(transform="compositional") %>% 
  phyloseq::psmelt() %>% 
  dplyr::select(Genus,Participante,Center,Abundance,Timepoint) %>% 
  dplyr::mutate(Genus=gsub("-",".",Genus)) %>% 
  dplyr::filter(Genus %in% sign_genus)


ps_df %>% 
  dplyr::filter(Genus=="Roseburia") %>% 
  dplyr::arrange(Participante,Center,Timepoint) %>% 
  ggstatsplot::grouped_ggwithinstats(Timepoint,Abundance,grouping.var = Center,type="robust")

ps_df %>% 
  dplyr::group_by(Participante,Genus) %>% 
  dplyr::arrange(Participante,Genus,Timepoint) %>% 
  tidyr::pivot_wider(names_from = Timepoint,values_from = Abundance) %>% 
  dplyr::mutate(Delta=Month6-BL) %>% 
  dplyr::select(-BL,-Month6) %>% 
  ggstatsplot::grouped_ggbetweenstats(x=Genus,y=Delta,grouping.var = Center)

ps_df %>% 
  dplyr::group_by(Participante,Genus) %>% 
  dplyr::arrange(Participante,Genus,Timepoint) %>% 
  tidyr::pivot_wider(names_from = Timepoint,values_from = Abundance) %>% 
  dplyr::mutate(Delta=Month6-BL) %>% 
  dplyr::select(-BL,-Month6) %>% 
  tidyr::pivot_wider(names_from = Genus,values_from = Delta) %>% 
  dplyr::ungroup() %>% 
  dplyr::select(-Participante,-Center) %>% 
  as.matrix() %>% cor() %>% 
  corrplot::corrplot()
  
  
```

## Core Microbiome

### phyloseqGraphTest

This is an alternative to PERMANOVA, based on distance-based graph neighbouring and permutation.

```{r core microbiome chunk1, include=TRUE, echo=F, message=FALSE, warning=FALSE, paged.print=FALSE, tidy=T,eval=F}
suppressMessages(require(dplyr))
suppressMessages(require(phyloseq))
suppressMessages(require(ggplot2))
suppressMessages(require(patchwork))

set.seed(1234)
### Species Level

devtools::load_all("~/Documents/Work/Development/WMGSPipeline/")
myMREObject<-readRDS("data/cfu/mreObject.rds")

my_ps<-myMREObject %>% metar::get_phyloseq(.,type="dada2") %>% 
  phyloseq::subset_samples(Center=="C01") %>% 
  phyloseq::tax_glom(.,taxrank="Genus")

gt = phyloseqGraphTest::graph_perm_test(my_ps, sampletype = "Timepoint",
                     distance = "bray", type = "mst",)

phyloseqGraphTest::plot_test_network(gt)
phyloseqGraphTest::plot_permutations(gt)

myMREObject %>% metar::get_phyloseq(.,type="dada2") %>% 
  phyloseq::subset_samples(Center=="C02") %>% 
  phyloseq::tax_glom(.,taxrank="Genus") %>% 
  phyloseqGraphTest::graph_perm_test(., sampletype = "Timepoint",
                     distance = "bray", type = "mst") %>% 
  phyloseqGraphTest::plot_permutations(.)

```
### Spiec-Easi

This packages uses inverse covariance matrices to build ecological networks. In this analysis we aim to see:

* Whether the ecological network  is different, at baseline, for the CG and IG. It should not be.
* Whether there is any alteration in the ecological network of CG and IG separately, at Month6

#### Baseline

```{r core microbiome chunk2 BL, include=TRUE, echo=F, message=FALSE, warning=FALSE, paged.print=FALSE, tidy=T,eval=T}
suppressMessages(require(dplyr))
suppressMessages(require(phyloseq))
suppressMessages(require(ggplot2))
suppressMessages(require(patchwork))
suppressMessages(require(SpiecEasi))
set.seed(1234)
### Species Level

devtools::load_all("~/Documents/Work/Development/WMGSPipeline/")
myMREObject<-readRDS("data/cfu/mreObject.rds")

### IG/C01
set.seed(1234)
my_ps_C01_BL<-myMREObject %>% metar::get_phyloseq(.,type="dada2") %>% 
  phyloseq::rarefy_even_depth() %>% 
  microbiome::core(prevalence = 0.3,detection=2) %>%
  phyloseq::tax_glom(.,taxrank="Genus") %>% 
  phyloseq::subset_samples(Center=="C01") %>%
  phyloseq::subset_samples(Timepoint=="BL") 


### Build network and initial graph
se_mb_C01_BL <- spiec.easi(my_ps_C01_BL, method='mb', lambda.min.ratio=1e-2,
                           nlambda=20, pulsar.params=list(rep.num=50,ncores=4))
ig2_mb_C01_BL <- adj2igraph(getRefit(se_mb_C01_BL),  vertex.attr=list(name=taxa_names(my_ps_C01_BL)))
ig2_mb_C01_BL_LO = igraph::layout_with_fr(ig2_mb_C01_BL)

### prune isolated nodes
Isolated = which(igraph::degree(ig2_mb_C01_BL)==0)
G2 = igraph::delete.vertices(ig2_mb_C01_BL,Isolated)
LO2 = ig2_mb_C01_BL_LO[-Isolated,]

## Plot Network
plot_network(G2, my_ps_C01_BL, type='taxa', color="Family",)



### CG/C02
set.seed(1234)
my_ps_C02_BL<-myMREObject %>% metar::get_phyloseq(.,type="dada2") %>% 
  phyloseq::rarefy_even_depth() %>% 
  microbiome::core(prevalence = 0.3,detection=2) %>%
  phyloseq::tax_glom(.,taxrank="Genus") %>% 
  phyloseq::subset_samples(Center=="C02") %>%
  phyloseq::subset_samples(Timepoint=="BL") 

### Build network and initial graph
se_mb_C02_BL <- spiec.easi(my_ps_C02_BL, method='mb', lambda.min.ratio=1e-2,
                           nlambda=20, pulsar.params=list(rep.num=50,ncores=4))
ig2_mb_C02_BL <- adj2igraph(getRefit(se_mb_C02_BL),  vertex.attr=list(name=taxa_names(my_ps_C02_BL)))
ig2_mb_C02_BL_LO = igraph::layout_with_fr(ig2_mb_C02_BL)

### prune isolated nodes
Isolated = which(igraph::degree(ig2_mb_C02_BL)==0)
G2 = igraph::delete.vertices(ig2_mb_C02_BL,Isolated)
LO2 = ig2_mb_C02_BL_LO[-Isolated,]

## Plot Network
plot_network(G2, my_ps_C02_BL, type='taxa', color="Family",)

```
#### Month6

```{r core microbiome chunk3 Month6, include=TRUE, echo=F, message=FALSE, warning=FALSE, paged.print=FALSE, tidy=T,eval=T}
suppressMessages(require(dplyr))
suppressMessages(require(phyloseq))
suppressMessages(require(ggplot2))
suppressMessages(require(patchwork))
suppressMessages(require(SpiecEasi))
set.seed(1234)
### Species Level

devtools::load_all("~/Documents/Work/Development/WMGSPipeline/")
myMREObject<-readRDS("data/cfu/mreObject.rds")

### IG/C01
set.seed(1234)
my_ps_C01_Month6<-myMREObject %>% metar::get_phyloseq(.,type="dada2") %>% 
  phyloseq::rarefy_even_depth() %>% 
  microbiome::core(prevalence = 0.3,detection=2) %>%
  phyloseq::tax_glom(.,taxrank="Genus") %>% 
  phyloseq::subset_samples(Center=="C01") %>%
  phyloseq::subset_samples(Timepoint=="Month6") 

### Build network and initial graph
se_mb_C01_Month6 <- spiec.easi(my_ps_C01_Month6, method='mb', lambda.min.ratio=1e-2,
                           nlambda=20, pulsar.params=list(rep.num=50,ncores=4))
ig2_mb_C01_Month6 <- adj2igraph(getRefit(se_mb_C01_Month6),  vertex.attr=list(name=taxa_names(my_ps_C01_Month6)))
ig2_mb_C01_Month6_LO = igraph::layout_with_fr(ig2_mb_C01_Month6)

### prune isolated nodes
Isolated = which(igraph::degree(ig2_mb_C01_Month6)==0)
G2 = igraph::delete.vertices(ig2_mb_C01_Month6,Isolated)
LO2 = ig2_mb_C01_Month6_LO[-Isolated,]

## Plot Network
plot_network(G2, my_ps_C01_Month6, type='taxa', color="Family",)



### CG/C02
set.seed(1234)
my_ps_C02_Month6<-myMREObject %>% metar::get_phyloseq(.,type="dada2") %>% 
  phyloseq::rarefy_even_depth() %>% 
  microbiome::core(prevalence = 0.3,detection=2) %>%
  phyloseq::tax_glom(.,taxrank="Genus") %>% 
  phyloseq::subset_samples(Center=="C02") %>%
  phyloseq::subset_samples(Timepoint=="Month6") 

### Build network and initial graph
se_mb_C02_Month6 <- spiec.easi(my_ps_C02_Month6, method='mb', lambda.min.ratio=1e-2,
                           nlambda=20, pulsar.params=list(rep.num=50,ncores=4))
ig2_mb_C02_Month6 <- adj2igraph(getRefit(se_mb_C02_Month6),  vertex.attr=list(name=taxa_names(my_ps_C02_Month6)))
ig2_mb_C02_Month6_LO = igraph::layout_with_fr(ig2_mb_C02_Month6)

### prune isolated nodes
Isolated = which(igraph::degree(ig2_mb_C02_Month6)==0)
G2 = igraph::delete.vertices(ig2_mb_C02_Month6,Isolated)
LO2 = ig2_mb_C02_Month6_LO[-Isolated,]

## Plot Network
plot_network(G2, my_ps_C02_Month6, type='taxa', color="Family",)

```
