---
site: workflowr::wflow_site
title: "Microbiome association with Mediterranean diet"
author: "Marc Noguera-Julian, PhD. "
date: "`r format(Sys.time(), '%Y-%m-%d')`"
header-includes:
- \usepackage{fancyhdr}
- \usepackage{lipsum}
- \pagestyle{fancy}
- \fancyhead[CO,CE]{Microbioma intestinal en intervencion nutricional}
- \fancyfoot[LO,RE]{Marc Noguera-Julian, IrsiCaixa}
- \fancyhead[LE,RO]{\thepage}
- \fancyfoot[CO,CE]{}
- \usepackage[default]{sourcesanspro}
- \usepackage[T1]{fontenc}
- \usepackage{eso-pic,graphicx,transparent}
mainfont: Helvetica
output: html_document
---

```{r setup, include=TRUE, echo=F, message=FALSE, warning=FALSE, paged.print=FALSE, tidy=F}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_knit$set(root.dir = here::here())
rmd_env <- "devel"
```

## Temporal Analysis

Since MedCars is a longitudinal study, we need to be awar of temporal trends that may be affecting our results.

We can see that there is a clear difference in the timing of visits in different centers. CG seems to have planned visits later on (12 days later, median difference). Can this affect results?

```{r follow-up chunk1, include=TRUE, echo=T, message=FALSE, warning=FALSE, paged.print=FALSE, tidy=T}
suppressWarnings(require(phyloseq))
suppressWarnings(require(phylosmith))
suppressWarnings(require(patchwork))
suppressWarnings(require(vegan))
suppressWarnings(require(ggstatsplot))
suppressWarnings(require(xlsx))
# Load package from local copy
suppressWarnings(devtools::load_all("~/Documents/Work/Development/WMGSPipeline/"))
ps_silva<-readRDS("data/cfu/DADA2/ps_silva.rds")
myMREObject<-readRDS("data/cfu/mreObject.rds")
myMREObject@wd<-"output/mre_data/cfu"

metadata_df<-get_meta(myMREObject)

metadata_df<-metadata_df %>% 
  dplyr::mutate(visitDate=as.Date(visitDate,format = "%m/%d/%y"))

# metadata_df %>%
#   ggplot(.,aes(x=visitDate,y=Participante,group=Participante,color=Center))+geom_point()+geom_line()

metadata_df %>% 
  dplyr::filter(Timepoint=="BL") %>% 
  dplyr::mutate(days=lubridate::yday(visitDate)) %>%
  ggstatsplot::ggbetweenstats(.,y = days,x=Center,title = "Date of visit")
```

### Time of Visit and microbiome

Let's see if there is a major trend in visit date vs microbiome composition. There is no apparent trend depending on time og visit. That's reassuring.

```{r follow-up chunk 2, include=TRUE, echo=T, message=FALSE, warning=FALSE, paged.print=FALSE, tidy=T}
phylum_df<-metar::get_phyloseq(myMREObject,type="dada2") %>% 
  microbiome::transform(.,transform="compositional") %>% 
  phyloseq::tax_glom(.,taxrank="Phylum",NArm=F) %>% 
  phyloseq::psmelt() %>% 
  dplyr::left_join(metadata_df %>%
                     dplyr::select(c("Center","Participante","Timepoint","visitDate")),by=c("Center","Participante","Timepoint","visitDate")) %>%
  dplyr::filter(Timepoint=="BL")

phylums<-phylum_df$Phylum %>% unique()

phylum_df %>% 
  dplyr::group_by(Phylum,visitDate,Center) %>% 
  dplyr::summarise(Abundance=mean(Abundance)) %>% 
  ggplot(.,aes(x=visitDate,y=Abundance,fill=Phylum))+geom_area()+facet_grid(~Center)

phylum_df %>% 
  dplyr::filter(Phylum == "Firmicutes") %>% 
  ggstatsplot::ggscatterstats(visitDate,Abundance,type = "robust")
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
